---
title: "Summarize samples"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, echo = F, message=F, warning=F}
library(tidyverse)
library(knitr)
library(skimr)
```


### Load data

```{r load data, message = F, warning = F}
### This is the full primary sample log
sample_log <- read_csv('FLC Primary Sample Log.csv') %>%
  separate(Sample, c("individual", "string"), sep = "_", remove = F)

### This is the ChROseq table from Tim's Cell Reports paper
chro_samples <- read_csv("ChROseq_samples.csv")
chro_samples <- substr(names(chro_samples)[-1], 5, 14)

### These are the RNAseq samples
rna <- read_tsv("pFLC01-34_CountTable_DESeq2Normalized_noTCGA_20200202.txt")
rna_samples = names(rna)[-1]
```

---

### Check which samples are in both RNAseq and ChROseq and which are unique

```{r overlap, message = F, warning = F}

### ChROseq and RNAseq samples
both_rna_chro <- intersect(chro_samples, rna_samples)
rna_only <- setdiff(rna_samples, chro_samples)
chro_only <- setdiff(chro_samples, rna_samples)
```

These are shared samples: **`r both_rna_chro`**

These are rna only: **`r rna_only`**

These are chro only: **`r chro_only`**

---

### Check for paired individual samples in the RNAseq & ChROseq sample set

```{r check both, message = F, warning = F}
both <- sample_log %>%
  filter(Sample %in% both_rna_chro)

table(both$individual, both$`Cancer Status`)  
```

For samples that have both ChROseq and RNAseq data, only **FLC09 (FLC09_MUOD & FLC_TZOG)** and **FLC27 (FLC27_XDGP & FLC27_BWSX)** have an NML and FLC sample.

---

### Check for paired individual samples in the RNAseq only set

```{r check rna only, message = F, warning = F}
rna_only_df <- sample_log %>%
  filter(Sample %in% rna_only)

table(rna_only_df$individual, rna_only_df$`Cancer Status`)  
```

For samples that are only present in the RNAseq data, only **FLC06 (FLC06_GMKX, FLC06_DZIS, FLC06_UFNW, & FLC06_SLVP)** and **FLC34 (FLC34_YROP & FLC34_PMVV)** have an NML and FLC sample.

---

### Check for paired individual samples in the ChROseq only sample set

```{r check chro only, message = F, warning = F}
chro_only_df <- sample_log %>%
  filter(Sample %in% chro_only)

table(chro_only_df$individual, chro_only_df$`Cancer Status`)  
```

None of the ChROseq only samples have paired NML/FLC.

---

### Check for individual samples (FLC or NML) with both ChRO + RNA

```{r}
sample_matched.data <- sample_log %>% 
  select("Sample", "Cancer Status", "ChRO-seq", "ChROseq included in Cell Reports", "FLCs sequenced with (mRNA)", "mRNA included in Cell Reports") %>% 
  filter(`ChROseq included in Cell Reports` == "Y" & `mRNA included in Cell Reports` == "Y")
kable(sample_matched.data)
```
These are all the patient samples that contain both RNA-seq and ChRO-seq.
Of these, two sets are matched: **FLC09 and FLC27**.

---

```{r import real masses and dedup}
FLCmasses <- read_csv("FLCmasses.csv")
FLC_Primary_Sample_Log <- read_csv("FLC Primary Sample Log.csv")
```

```{r sequeneced and published}
seq.pub.rna <- sample_log %>% 
  select("Sample", "Cancer Status", "mRNA included in Cell Reports", "ChRO-seq", "ChROseq included in Cell Reports", "FLCs sequenced with (mRNA)") %>% 
  filter(`mRNA included in Cell Reports` == "Y")
no.dups <- FLC_Primary_Sample_Log %>% 
  select("Sample", "Remove Sample as Duplicate") %>% 
  filter(`Remove Sample as Duplicate` == "N")
seq.pub.combined <- inner_join(seq.pub.rna, no.dups)
kable(seq.pub.rna)
write.csv(x = seq.pub.rna, file = "output/seq_pub.csv")
```

```{r sequenced and not published}
seq.unpub.rna <- sample_log %>% 
  select("Sample", "Cancer Status", "FLCs sequenced with (mRNA)", "mRNA included in Cell Reports", "ChRO-seq", "ChROseq included in Cell Reports") %>% 
  filter(`mRNA included in Cell Reports` == "N" & !is.na(`FLCs sequenced with (mRNA)`))
seq.unpub.combined <- inner_join(seq.unpub.rna, FLCmasses)
kable(seq.unpub.rna)
write.csv(x = seq.unpub.combined, file = "output/seq_unpub.csv")
```

```{r unsequenced RNA}
unseq.rna <- sample_log %>% 
  select("Sample", "Cancer Status", "FLCs sequenced with (mRNA)", "mRNA included in Cell Reports", "ChRO-seq", "ChROseq included in Cell Reports") %>% 
  filter_all(any_vars(is.na(.)))
kable(unseq.rna)
unseq.combined <- inner_join(unseq.rna, FLCmasses)
write.csv(x = unseq.combined, file = "output/unseq_rna.csv")
```







