---
title: "scATAC Tutorial"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install and load libraries, , echo = F, message = F, warning = F}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
devtools::install_github("GreenleafLab/ArchR", ref="master", repos = BiocManager::repositories())
library(ArchR)
ArchR::installExtraPackages()
set.seed(1)
addArchRThreads(threads = 16) 
```

We need a reference genome for downstream analyses. ArchR natively supports hg19, hg38, mm9, and mm10.

```{r download tutorial data and add genome, eval=FALSE}
inputFiles <- getTutorialData("Hematopoiesis")
inputFiles
addArchRGenome("hg19")
```

### Creating Arrow Files

ArchR uses arrow files, the base unit of an ArchR analytical project. Every arrow file stores all the data associated with an individual sample (i.e. a single replicate of a particular condition). During creation and as additional analyses are performed, ArchR updates and edits each Arrow file to contain additional layers of information. IAn Arrow file is actually just a path to an external file stored on disk. More explicitly, an Arrow file is not an R-language object that is stored in memory but rather an HDF5-format file stored on disk.

```{r}
ArrowFiles <- createArrowFiles(
  inputFiles = inputFiles,
  sampleNames = names(inputFiles),
  filterTSS = 4, #Dont set this too high because you can always increase later
  filterFrags = 1000, 
  addTileMat = TRUE,
  addGeneScoreMat = TRUE
)
#We can inspect the ArrowFiles object to see that it is actually just a character vector of Arrow file paths.
ArrowFiles
```

Quality Control of scATAC-seq data is essential to remove cells that contribute to low-quality data. There are three characteristics that scATAC considers 
1. The number of unique nuclear fragments (i.e. not mapping to mitochondrial DNA).
2. The signal-to-background ratio. Low signal-to-background ratio is often attributed to dead or dying cells which have de-chromatinzed DNA which allows for random transposition genome-wide.
3. The fragment size distribution. Due to nucleosomal periodicity, we expect to see depletion of fragments that are the length of DNA wrapped around a nucleosome (approximately 147 bp).

ArchR also infers doublets, a single droplet that contains multiple cells.
```{r infer doublets}
doubScores <- addDoubletScores(
  input = ArrowFiles,
  k = 10, #Refers to how many cells near a "pseudo-doublet" to count.
  knnMethod = "UMAP", #Refers to the embedding to use for nearest neighbor search.
  LSIMethod = 1
)
```

###Creating an ArchRProject
```{r Create ArchRProject and filter}
proj <- ArchRProject(
  ArrowFiles = ArrowFiles, 
  outputDirectory = "HemeTutorial",
  copyArrows = TRUE #This is recommened so that you maintain an unaltered copy for later usage.
)

#We can query which data matrices are available in the ArchRProject. At this point in time, we should have “GeneScoreMatrix” and “TileMatrix”. As we continue to work and add to the ArchRProject, we can use the following function to query which matricies are added to the project.
getAvailableMatrices(proj)

#Next we can filter out putative doublets based on the scores established in the `infer doublets` chunk. Importantly, this does not delete the data from the Arrow files, but rather forces ArchRProject to ignore these cells. 
proj <- filterDoublets(ArchRProj = proj)
```

###Dimensionality Reduction and Clustering

```{r Dimensionality Reduction, UMAP Generation and Visualization}

proj <- addIterativeLSI(ArchRProj = proj, useMatrix = "TileMatrix", name = "IterativeLSI")
proj <- addClusters(input = proj, reducedDims = "IterativeLSI")
proj <- addUMAP(ArchRProj = proj, reducedDims = "IterativeLSI")
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample", embedding = "UMAP")
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Clusters", embedding = "UMAP")
ggAlignPlots(p1, p2, type = "h")
plotPDF(p1,p2, name = "Plot-UMAP-Sample-Clusters.pdf",
        ArchRProj = proj, addDOC = FALSE, width = 5, height = 5)
```









